NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-10 Manuscript/manuscriptPS.tex
Update text

Binding of coions to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of ion concentration \cite{catte16}. This is less straighforward
for charged lipid bilayers because counterions are always present and the
ion free reference state does not exist
(section \ref{mixtureTOadditionalCIs} in the supplementary information).
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

>>>>

Binding of coions to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of ion concentration \cite{catte16}. This is less straighforward
for charged lipid bilayers because counterions are always present and the
ion free reference state does not exist
(section \ref{mixtureTOadditionalCIs} in the supplementary information).
Here we evaluate the amount of bound charge using the changes of
headgroup order parameters with increasing amount of negatively charged lipids in the bilayer.
According to the electrometer concept, the headgroup order parameters of POPC
increase when negativaly charged POPS lipids are incorporated in lipid bilayer
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions having
the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}),
while other simulations predict no change or decrease in the order parameters
(Fig. \ref{HGorderparametersPCvsPS}).
%In experiments and MacRog simulations, the increasing amount of negative charge
%in the membrane tilts the PC headgroup more parallel to the membrane plane
%leading to the larger order parameters.
In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS.
Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}
